Retrofitting Effect Handlers onto OCaml

### 2020-10-30 retro-concurrency.tex
Minor fix
Added static semantics

\lstset{
			language=caml,
      basicstyle=\ttfamily\footnotesize,
      flexiblecolumns=false,
			tabsize=2,
			escapechar={<@}{@>},
      %basewidth={0.5em,0.45em},
      %aboveskip={3pt},
      %belowskip={3pt},
			breaklines = true,
      breakatwhitespace = true,
      keywordstyle=\color{Bittersweet}\bfseries,
      commentstyle=\color{blue}\itshape,
      stringstyle=\color{MidnightBlue},
			keywords=[2]{},
			keywordstyle=[2]\color{byzantine}\bfseries,
			keywords=[3]{effect,perform,continue,discontinue,clone_continuation},
			keywordstyle=[3]\color{caribbeangreen}\bfseries,
			keywords=[4]{call,return,external},
			keywordstyle=[4]\color{Bittersweet}\bfseries,
			classoffset=1,
			upquote=true,
			keywordstyle=\color{byzantine}\bfseries,
			classoffset=0,
			mathescape=true,
			numberstyle=\tiny\color{gray},
			numbersep=5pt
}

@@

\lstMakeShortInline[columns=fullflexible]|

@@

\begin{enumerate}[label=R\arabic*]
	\item \textbf{Backwards compatibility} A well behaved OCaml program does not
		break under OCaml extended with effect handlers. OCaml code that does not
		use effect handlers will pay negligible performance cost.
	\item \textbf{Tool compatibility} OCaml with effect handlers produce
		well-formed backtraces, and remains compatible with program analysis tools
		such as debuggers and profilers that inspect the stack using DWARF unwind
		tables.
	\item \textbf{Effect handler efficiency} In order to support
		highly-concurrent applications, the program must accommodate millions of
		continuations in the heap. Installing effect handlers, capturing and
		resuming continuations must be fast.
\if{0}
	\item \textbf{Effect handler expressivity} While making continuations
		resumable at-most once (\emph{affine}) is a standard way to make
		continuation capture fast, we must still support use cases where the
		continuations are resumed more than once.
\fi
	\item \textbf{Forwards compatibility} While this work does not present an
		effect system, the design must be compatible with a future effect system
		for OCaml. As a cornerstone of modularity, we also want blocking I/O code
		to \emph{transparently} be made asynchronous with the help of effect
		handlers.
\end{enumerate}

@@

The same backtrace can also be obtained by using \emph{frame pointers} instead
of DWARF unwind tables. OCaml allows compiling code with frame pointers, but
they are not enabled by default. OCaml stack tends to be deep with small frames
due to the pervasive use of recursive functions, not all of which is
tail-recursive. Hence, the addition of frame pointers can significantly
increase the size of the
stack~\footnote{https://github.com/ocaml/ocaml/issues/5721\#issuecomment-472965549}.
Moreover, not using frame pointers saves two instructions in the function
prologue and epilogue, and makes an extra register (|rbp| on x86) available.

@@

\if{0}
\subsection{Multi-shot continuations}

Though our primary goal is to enable non-local control flow mechanisms, effect
handlers have been gaining increasing attention for Bayesian machine learning
and probabilisitic programming~\cite{PyroJMLR,MooreEdward2Google}, which resume
the continuations more than once for performing inference over the
probabilistic program. To enable these use cases, we provide a
|clone_continuation| primitive that makes a copy of the given continuation.
Care must be taken when using multi-shot continuations as they do not mix well
with code that uses resources. With multi-shot continuations, a function
\emph{may return more than once}. Consider the code below:

\begin{lstlisting}
let foo m = ...;perform E; ...
let bar () =
  let m = malloc 64 in
	begin try foo m with _ -> () end;
	free m
let _ = match bar () with
| () -> ()
| effect E k ->
    let k' = clone_continuation k in
		continue k (); continue k' ()
\end{lstlisting}

The function |bar| expects |foo| to return exactly once, ignoring any exceptions. However,
due to continuation cloning, the function returns twice, leading to double-free
error.
\fi

@@

>>>>

\lstset{
			language=caml,
      basicstyle=\ttfamily\footnotesize,
      flexiblecolumns=false,
			tabsize=2,
			escapechar={<@}{@>},
      %basewidth={0.5em,0.45em},
      %aboveskip={3pt},
      %belowskip={3pt},
			breaklines = true,
      breakatwhitespace = true,
      keywordstyle=\color{Bittersweet}\bfseries,
      commentstyle=\color{blue}\itshape,
      stringstyle=\color{MidnightBlue},
			keywords=[2]{},
			keywordstyle=[2]\color{byzantine}\bfseries,
			keywords=[3]{eff,effect,perform,continue,discontinue,continuation,
			             clone_continuation},
			keywordstyle=[3]\color{caribbeangreen}\bfseries,
			keywords=[4]{call,return,external},
			keywordstyle=[4]\color{Bittersweet}\bfseries,
			classoffset=1,
			upquote=true,
			keywordstyle=\color{byzantine}\bfseries,
			classoffset=0,
			mathescape=true,
			numberstyle=\tiny\color{gray},
			numbersep=5pt
}

@@

\lstMakeShortInline[columns=fullflexible,keepspaces]|

@@

\begin{enumerate}[label=R\arabic*]
	\item \textbf{Backwards compatibility} A well behaved OCaml program does not
		break under OCaml extended with effect handlers. OCaml code that does not
		use effect handlers will pay negligible performance cost.
	\item \textbf{Tool compatibility} OCaml with effect handlers produce
		well-formed backtraces, and remains compatible with program analysis tools
		such as debuggers and profilers that inspect the stack using DWARF unwind
		tables.
	\item \textbf{Effect handler efficiency} In order to support
		highly-concurrent applications, the program must accommodate millions of
		continuations in the heap. Installing effect handlers, capturing and
		resuming continuations must be fast.
	\item \textbf{Forwards compatibility} While this work does not present an
		effect system, the design must be compatible with a future effect system
		for OCaml. As a cornerstone of modularity, we also want blocking I/O code
		to \emph{transparently} be made asynchronous with the help of effect
		handlers.
\end{enumerate}

@@

The same backtrace can also be obtained by using \emph{frame pointers} instead
of DWARF unwind tables. OCaml allows compiling code with frame pointers, but
they are not enabled by default. OCaml stack tends to be deep with small frames
due to the pervasive use of recursive functions, not all of which are
tail-recursive. Hence, the addition of frame pointers can significantly
increase the size of the
stack~\footnote{https://github.com/ocaml/ocaml/issues/5721\#issuecomment-472965549}.
Moreover, not using frame pointers saves two instructions in the function
prologue and epilogue, and makes an extra register (|rbp| on x86) available.

@@

@@

In this section, we formalise the effect handler design in Multicore OCaml.

\subsection{Static semantics}

As mentioned earlier, effect handlers in Multicore OCaml do not guarantee
effect safety, but only guarantee type safety. Programs without matching effect
handlers are well-typed Multicore OCaml programs. As a result, our static
semantics is simpler than languages that ensure effect
safety~\cite{Links,Koka,Effekt,Frank,HeliumBiernacki}. This is important for
backwards compatibility as our goal is to retrofit effect handlers to a
language with large legacy codebases. Programs that do not use effects remain
well-typed, and those that do compose well with those that don't.

The static semantics of effect handlers in OCaml is captured succinctly by its
API:
\begin{lstlisting}
type 'a eff
type ('a,'b) continuation
val perform: 'a eff -> 'a
val continue: ('a,'b) continuation -> 'a -> 'b
val discontinue: ('a,'b) continuation -> exn -> 'b

type 'a computation = unit -> 'a
type ('a,'b) handler =
{ret: 'a -> 'b;
 eff_1: 'c1.'c1 eff -> ('c1,'b)continuation -> 'b;
 ...
 eff_n: 'cn.'cn eff -> ('cn,'b)continuation -> 'b}
val match_with: 'a computation -> ('a,'b) handler
                               -> 'b
\end{lstlisting}

We introduce the type |'a eff| of effect values which when performed using the
|perform| primitive returns a |'a| value. Constructors for the value of type
|'a eff| are declared using the effect declarations. For example, with the
declaration |effect E : string -> int|, the effect value |E "42"| has the type
|int eff|. We introduce the type |('a,'b) continuation| of delimited
continuations which expect a |'a| value for resumption and return |'b| value.
The continuations may be |continue|d with a suitably typed value or
|discontinue|d with an exception.

For handling the effects, our implementation extends OCaml's |match ... with|
syntax with effect patterns. For the sake of exposition, we introduce a
|('a,'b) handler| type. This handler handles a |'a computation| that returns a
|'a| value, and itself returns a |'b| value. The handler has a single return
case of type |'a -> 'b|, and one or more effect cases. Each effect case,
handles a effect of type |'c eff|, whose continuation is of type
|('c,'b) continuation|, and returns a value of type |'b|. For example, in the
following code:
\begin{lstlisting}
effect E : string -> int
effect F : float -> int
let r = match perform (E "42") with
| v -> v == 0
| effect (E s) k ->
    not (continue k (int_of_string s))
| effect (F f) k ->
    not (continue k (int_of_float f))
\end{lstlisting}
\noindent |r| has type |bool|.

\subsection{Dynamic semantics}
